---
title: "Crosstable interface to MS-Word"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{crosstable-report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(crosstable)
library(dplyr)
```

## Create reports with `officer`

The real power of `crosstable` comes out when used with David Gohel's [awesome package ](https://davidgohel.github.io/officer/) `officer`, which allows to automatically create MS Word documents. 

Therefore, `crosstable` has several functions to output a beautiful report:

```{r officer, message=FALSE, warning=FALSE}
library(officer)
ct1=crosstable(iris, by=Species, test=TRUE)
ct2=crosstable(mtcars2, mpg,cyl,disp, by=am, effect=TRUE, total="both", showNA="always")
ct3=crosstable(esoph)

doc = read_docx() %>% 
  body_add_title("Dataset iris", 1) %>%
  body_add_title("Not compacted", 2) %>%
  body_add_normal("Automatic testing is bad and you should feel bad.") %>%
  body_add_crosstable(ct1) %>%
  body_add_title("Compacted", 2) %>%
  body_add_normal("When compacting, you might want to remove the test names.") %>%
  body_add_crosstable(ct1, compact=TRUE, show.test.name=FALSE) %>%
  body_add_break %>%
  body_add_title("Dataset mtcars2", 1) %>%
  body_add_glued("This dataset has {nrow(ct3)} rows and {x} columns.", x=ncol(ct3)) %>%
  body_add_normal("Look, there are labels!") %>%
  body_add_crosstable(ct2, compact=TRUE) %>%
  body_add_break %>%
  body_add_title("Dataset esoph", 1) %>%
  body_add_normal("This one was compacted beforehand.") %>%
  body_add_crosstable(compact(ct3)) %>%
  body_add_break 

```

## Output

To output, just use the `print` function:

```{r print, eval=FALSE}
filename="..\\examples\\officer-vignette.docx" #`\\` are needed for shell.exec on Windows
print(doc, filename) #write the docx file
shell.exec(filename) #open the docx file (fails if it is already open)
```

You can check out the example [here](https://github.com/DanChaltiel/crosstable/examples/officer-vignette.docx).

Unfortunately, large tables can overflow the MS Word document width. In this case (for instance for `mtcars2` in the previous example), you have to manually go to `Table Tools > Layout > AutoFit > AutoFit Window` in the ribbon to correct the width. 

You can learn more on officer on [its documentation](https://davidgohel.github.io/officer/).



## Create reports with `Rmarkdown`

<!-- Intégration Rmd in Rmd : la coloration syntaxique RStudio est mauvaise mais le code est bon ! -->
<!-- https://stackoverflow.com/questions/53226493/add-markdown-code-chunk-to-r-markdown-document#comment104859956_53226493 -->
Knitting (`knitr::knit()` ou via **RStudio**) this `Rmd` code also creates a MS-Word file. Here, you can use the power of `bookdown` to generate the automatic numbering of the tables.

````markdown

---
title: "Iris"
output: bookdown::word_document2
---
    
`r ''````{r setup, include=FALSE}
library(biostat2)
library(flextable)
library(dplyr) #pour le pipe %>% 
```

Table iris is given in Table \@ref(tab:irisTable).

`r ''````{r description, echo=FALSE, results='asis'}
cat("<caption> (\\#tab:irisTable) Table Iris </caption> \n\r ")
cross(cbind(Sepal.Length, Sepal.Width) ~ Species, iris, test = TRUE, total="column") %>% cross_to_flextable %>% autofit
```

````

You can example files here: [cross_markdown.Rmd](examples/cross_markdown.Rmd) and [cross_markdown.docx](examples/cross_markdown.docx).

