---
title: "Make automatic reports"
author: "Dan Chaltiel"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: 
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Make automatic reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(crosstable)
library(dplyr)
```

## Create reports with `officer`

The real power of `crosstable` comes out when used with David Gohel's awesome package [`officer` ](https://davidgohel.github.io/officer/), which allows to automatically create MS Word documents. 

Therefore, `crosstable` has several helper functions to output a beautiful report:

```{r officer, message=FALSE, warning=FALSE}
library(officer)
library(ggplot2)
ct1=crosstable(iris, by=Species, test=TRUE)
ct2=crosstable(mtcars2, c(mpg,cyl,disp), by=am, effect=TRUE, total="both", showNA="always")
ct3=crosstable(esoph)
options(crosstable_units="cm")

my_plot = ggplot(data = iris ) +
  geom_point(mapping = aes(Sepal.Length, Petal.Length))

doc = read_docx() %>% 
  body_add_title("Dataset iris", 1) %>%
  body_add_title("Not compacted", 2) %>%
  body_add_normal("Table \\@ref(table_autotest) is an example. However, automatic testing is bad and I should feel bad.") %>%
  body_add_crosstable(ct1) %>%
  body_add_table_legend("Automatic testing is bad", bookmark="table_autotest") %>%
  body_add_normal("Let's add a figure as well. You can see in Figure \\@ref(fig_iris) that sepal length is somehow correlated with petal length.") %>%
  body_add_figure_legend("Relation between Petal length and Sepal length", bookmark="fig_iris") %>% 
  body_add_gg2(my_plot, w=14, h=10, scale=1.5) %>% 
  body_add_title("Compacted", 2) %>%
  body_add_normal("When compacting, you might want to remove the test names.") %>%
  body_add_crosstable(ct1, compact=TRUE, show_test_name=FALSE) %>%
  body_add_break() %>%
  body_add_title("Dataset mtcars2", 1) %>%
  body_add_normal("This dataset has {nrow(ct3)} rows and {x} columns.", x=ncol(ct3)) %>%
  body_add_normal("Look, there are labels!") %>%
  body_add_crosstable(ct2, compact=TRUE) %>%
  body_add_break() %>%
  body_add_title("Dataset esoph", 1) %>%
  body_add_normal("This one was compacted beforehand for some reason.") %>%
  body_add_crosstable(compact(ct3))
```

Beware when using the whole `tidyverse`, as `crosstable::compact` conflicts with `purrr::compact`. You can solve this using an alias (`my_compact=crosstable::compact`) or using the `conflicted` package.

```{r save, include=FALSE}
# stop("Working directory = ", getwd())
if(file.exists("../examples"))
  print(doc, "../examples/vignette_officer.docx")
```

## Output

To save the file, just use the `print` function:

```{r print, eval=FALSE}
filename=file.path("..", "examples", "vignette_officer.docx", fsep="\\")#`\\` is needed for shell.exec on Windows
print(doc, filename) #write the docx file
shell.exec(filename) #open the docx file (fails if it is already open)
```

You can check out this example [here](https://github.com/DanChaltiel/crosstable/raw/master/examples/vignette_officer.docx).

Unfortunately, large tables can overflow the MS Word document width. In this case (for instance for `mtcars2` in the previous example), you have to manually go to `Table Tools > Layout > AutoFit > AutoFit Window` in the ribbon to correct the width. 

You can learn more on officer on [its documentation](https://davidgohel.github.io/officer/).



## Create reports with `Rmarkdown`

<!-- Intégration Rmd in Rmd: la coloration syntaxique RStudio est mauvaise mais le code est bon ! -->
<!-- https://stackoverflow.com/questions/53226493/add-markdown-code-chunk-to-r-markdown-document#comment104859956_53226493 -->
Knitting (`knitr::knit()` or via **RStudio**) this `Rmd` code also creates a MS-Word file. Here, you can use the power of `bookdown` to generate the automatic numbering of the tables.

````markdown

---
title: "Iris"
output: bookdown::word_document2
---
    
`r ''````{r setup, include=FALSE}
library(crosstable)
library(flextable)
library(dplyr) #pour le pipe %>% 
```

Table iris is given in Table \@ref(tab:irisTable).

`r ''````{r description, echo=FALSE, results='asis'}
cat("<caption> (\\#tab:irisTable) Table Iris </caption> \n\r ")
crosstable(iris, Sepal.Length, Sepal.Width, by=Species, test = TRUE, total="column") %>% as_flextable
```

````

You can example files here: [vignette_markdown.Rmd](../examples/vignette_markdown.Rmd) and [vignette_markdown.docx](../examples/vignette_markdown.docx).

