---
title: "Make automatic reports"
author: "Dan Chaltiel"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: 
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Make automatic reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(crosstable)
library(dplyr)
```

## Create reports with `officer`

The real power of `crosstable` comes out when used with David Gohel's awesome package [`officer` ](https://davidgohel.github.io/officer/), which allows to automatically create MS Word documents. 

Therefore, `crosstable` exports several helper functions to easily output a beautiful report:

```{r officer, message=FALSE, warning=FALSE}
library(officer)
library(ggplot2)
ct1=crosstable(iris, by=Species, test=TRUE)
ct2=crosstable(mtcars2, c(mpg,cyl,disp), by=am, effect=TRUE, total="both", showNA="always")
ct3=crosstable(esoph)
options(crosstable_units="cm")

my_plot = ggplot(data = iris ) +
  geom_point(mapping = aes(Sepal.Length, Petal.Length))

doc = read_docx() %>% 
  body_add_title("Dataset iris", 1) %>%
  body_add_title("Not compacted", 2) %>%
  body_add_normal("Table \\@ref(table_autotest) is an example. However, automatic testing is bad and I should feel bad.") %>%
  body_add_crosstable(ct1) %>%
  body_add_table_legend("Automatic testing is bad", bookmark="table_autotest") %>%
  body_add_normal("Let's add a figure as well. You can see in Figure \\@ref(fig_iris) that sepal length is somehow correlated with petal length.") %>%
  body_add_figure_legend("Relation between Petal length and Sepal length", bookmark="fig_iris") %>% 
  body_add_gg2(my_plot, w=14, h=10, scale=1.5) %>% 
  body_add_title("Compacted", 2) %>%
  body_add_normal("When compacting, you might want to remove the test names.") %>%
  body_add_crosstable(ct1, compact=TRUE, show_test_name=FALSE) %>%
  body_add_break() %>%
  body_add_title("Dataset mtcars2", 1) %>%
  body_add_normal("This dataset has {nrow(ct3)} rows and {x} columns.", x=ncol(ct3)) %>%
  body_add_normal("Look, there are labels!") %>%
  body_add_crosstable(ct2, compact=TRUE) %>%
  body_add_break() %>%
  body_add_title("Dataset esoph", 1) %>%
  body_add_normal("This one was compacted beforehand for some reason.") %>%
  body_add_crosstable(compact(ct3)) #beware of any conflict with purrr::compact
```

#### Output

```{r save, include=FALSE}
# stop("Working directory = ", getwd())
if(file.exists("../examples"))
  print(doc, "../examples/vignette_officer.docx")
```

To see the resulting Word document, use `write_and_open()`. 

If you only want to save the document without opening it, simply use `print()`:

```{r print, eval=FALSE}
write_and_open(doc) #save and open the docx file in a temporary file
write_and_open(doc, "my_report.docx") #fails if it is already open
print(doc, "my_report.docx") #only save the docx file
```

You can check out the result of the example above [here](https://github.com/DanChaltiel/crosstable/raw/master/examples/vignette_officer.docx).

#### Functions

Here is a brief description of the functions used in this example:

 + `officer::read_docx()`: creates a bare MS Word document
 + `body_add_title()`: adds a title paragraph of any level
 + `body_add_normal()`: adds a normal style paragraph. You can also incorporate variables using the syntax `{{nrow(ct3)}}` and references using the syntax `\\@ref(my_bookmark)`.
 + `body_add_crosstable()`: adds a crosstable
 + `body_add_figure_legend()` and `body_add_table_legend()`: adds a figure/table legend. The `bookmark` is the key that can be added elsewhere in `body_add_normal()`.
 + `body_add_gg2()`: adds a ggplot. Unlike `officer::body_add_gg()`, you can change the unit using the `units` argument or the options `options(crosstable_units="cm")`.

Browse https://davidgohel.github.io/officer for more insight about how you can use `officer`.

#### Styles issues

Crosstables need [Word styles](https://support.microsoft.com/fr-fr/office/personnaliser-ou-cr%C3%A9er-des-styles-d38d6e47-f6fc-48eb-a607-1eb120dec563) to work properly, and default values are contained in the default template of `officer::read_docx()`. However, if you use a custom template inside `read_docx()`, several functions in `crosstable` might need **style** parameterization.

For instance, `body_add_figure_legend()` needs 2 styles, which are by default "Image Caption" and "strong". If these styles are not found in the template, an error will be raised. 

You can prevent this by using either parameters `body_add_figure_legend(legend_style="Figure Legend", style="Bold")` or options `options(crosstable_style_legend="Figure Legend", crosstable_style_strong="Bold")`.

See `?crosstable_options` for a list of all styles you can specify with options.

#### Post-production

+ References numbering is calculated by MS Word. Therefore, they do not exist yet when the document is written. To trigger the calculation, you need to manually update the references inside MS Word: select all (\kbd{Ctrl}+\kbd{A}), then update (\kbd{F9}). You might even need to do this several times. See `?body_add_table_legend` for more insight.

 + Large tables can sometimes overflow the MS Word document width. In this case (for instance for `mtcars2` in the previous example), you have to manually go to `Table Tools > Layout > AutoFit > AutoFit Window` in the ribbon to correct the width. This is a limitation that comes with `officer` ([link](https://ardata-fr.github.io/officeverse/faq.html#update-fields)).

You can learn more on officer on [its documentation](https://davidgohel.github.io/officer/).



## Create reports with `Rmarkdown`

<!-- Intégration Rmd in Rmd: la coloration syntaxique RStudio est mauvaise mais le code est bon ! -->
<!-- https://stackoverflow.com/questions/53226493/add-markdown-code-chunk-to-r-markdown-document#comment104859956_53226493 -->
Knitting (`knitr::knit()` or via **RStudio**) this `Rmd` code also creates a MS-Word file. Here, you can use the power of `bookdown` to generate the automatic numbering of the tables.

````markdown

---
title: "Iris"
output: bookdown::word_document2
---
    
`r ''````{r setup, include=FALSE}
library(crosstable)
library(flextable)
library(dplyr) #pour le pipe %>% 
```

Table iris is given in Table \@ref(tab:irisTable).

`r ''````{r description, echo=FALSE, results='asis'}
cat("<caption> (\\#tab:irisTable) Table Iris </caption> \n\r ")
crosstable(iris, Sepal.Length, Sepal.Width, by=Species, test = TRUE, total="column") %>% as_flextable
```

````

You can example files here: [vignette_markdown.Rmd](../examples/vignette_markdown.Rmd) and [vignette_markdown.docx](../examples/vignette_markdown.docx).

