---
title: "Introduction to Crosstable"
author: "Dan Chaltiel"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette: 
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{crosstable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r options, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=FALSE, 
  message=FALSE
  )
  options(width = 100)
  library(flextable)
```

```{r css, echo=FALSE, results="asis"}
cat("
<style>
kbd {
    background-color: #eee;
    border-radius: 3px;
    border: 1px solid #b4b4b4;
    box-shadow: 0 1px 1px rgba(0, 0, 0, .2), 0 2px 0 0 rgba(255, 255, 255, .7) inset;
    color: #333;
    display: inline-block;
    font-size: .85em;
    font-weight: 700;
    line-height: 1;
    padding: 2px 4px;
    white-space: nowrap;
}
</style>
")
```


## Crosstable

Crosstable is a package centered on a single function, `crosstable`, which easily computes descriptive statistics on datasets.

Before starting this vignette, here are a few points:

 + Tables look better on a white background. If you are using RStudio in dark mode, borders may look blurry.
 + If you are not familiar with `dplyr` and pipes (`%>%`, <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>M</kbd> in RStudio), I warmly recommend you to read Julien Barnier's [tidyverse tutorial](https://juba.github.io/tidyverse/10-dplyr.html#pipe). Nevertheless, this vignette should still be somehow understandable, since `crosstable` is perfectly usable with base `R`.


## Dataset: modified `mtcars`

First, since it uses the power of the `label` attribute, let's start by building a labelled dataset. In this vignette, we will use a modified version of the `mtcars` dataset, which comprises 11 aspects of design and performance for 32 automobiles. We modify it to add textual categories, make some numeric variables factors, and add labels from a table using `import_labels`. 

For convenience, this dataset is already packed into `crosstable` so don't bother re-creating it for your own tests.

```{r dataset}
library(crosstable)
library(dplyr)
mtcars_labels = read.table(header=TRUE, text="
  name label
  mpg  'Miles/(US) gallon'
  cyl  'Number of cylinders'
  disp 'Displacement (cu.in.)'
  hp   'Gross horsepower'
  drat 'Rear axle ratio'
  wt   'Weight (1000 lbs)'
  qsec '1/4 mile time'
  vs   'Engine'
  am   'Transmission'
  gear 'Number of forward gears'
  carb 'Number of carburetors'
")
mtcars2 = mtcars %>% 
  mutate(vs=ifelse(vs==0, "vshaped", "straight"),
         am=ifelse(am==0, "auto", "manual")) %>% 
  mutate_at(c("cyl", "gear"), factor) %>% 
  import_labels(mtcars_labels, name_from="name", label_from="label") 
#I also could have used `Hmisc::label` or `expss::apply_labels` to add labels
```

Here is the result when we apply `crosstable` on it:

```{r crosstable-bare}
crosstable(mtcars2) %>% head(11)
```
In this vignette, I only printed the `mpg`, `cyl` and `disp` variables (the 11 first rows), but all other variables were also described with this call. 

By default, numeric variables (like `mpg` and `disp`) are described with min/max, median/IQR, mean/sd and number of observations/missing, while categorical (factor/character) variables (like `cyl`) are described with levels counts and fraction. All of this is fully customizable, as you will see hereafter.




## Arguments

This is already pretty handy, but we  can easily add some control with :

+ unnamed arguments, to select the columns we want to describe
+ the `by` argument, to select the column we want to group on to calculate the stats.

For instance, if we wanted the statistics of `mpg` and `cyl` for both levels of `am`, we could write this:

```{r crosstable-args, tidy=TRUE}
crosstable(mtcars2, mpg, cyl, by=am)
```

The `by` column should be a factor, character or logical vector. If it is a numeric vector, then only numeric vectors can be described and correlation factors will be displayed.

There are many ways to select variables: with names, character vector, `tidyselect` helpers, formula... For instance, you could write `crosstable(mtcars2, starts_with("d"))`, `crosstable(mtcars2, is.numeric)`, `crosstable(mtcars2, -c(mpg:hp))`, or `crosstable(mtcars2, mpg~am)`. This is described in details in `vignette("crosstable-selection")` ([link](crosstable-selection.html)).




## HTML printing

This might be good enough for a preview, but printing a table to the console has never been the greatest thing to visualize tables, especially in a vignette like this one.

Fortunately, you can use the `as_flextable` function to turn any `crosstable` into a HTML, ready-to-print table, which will be automatically displayed in the Viewer if your are using RStudio. 


```{r crosstable-flextable}
crosstable(mtcars2, mpg, cyl, by=am) %>% 
  as_flextable(keep_id=TRUE)
```

If you want to share your tables in a MS Word document, you can use David Gohel's [awesome package `officer`](https://davidgohel.github.io/officer/). 

Here, I use the `keep_id` argument to keep the variable name, but in practice you usually drop it for the label. See `vignette("crosstable-officer")` ([link](crosstable-officer.html)) for more on `as_flextable` and on how to integrate crosstables with `officer` and `Rmarkdown`.


## Descriptive functions

There may be too much information with `crosstable`'s default descriptive function. In this case, you can use any set of function that fits your need. They will be applied to all numeric variables.

```{r crosstable-functions}
crosstable(mtcars2, am, mpg, wt, by=vs, funs=c(median, mean, sd)) %>% 
  as_flextable(keep_id=TRUE)
```

You might want to use `crosstable`'s convenience functions such as `moystd`, `mediqr`, `minmax`, or `nna`. For 
additional arguments, you can use `funs_arg` argument: `crosstable(mtcars3, c(disp, hp, am), by=vs, funs=c(moystd,quantile), funs_arg = list(dig=3, probs=c(0.25,0.75)))`. Numbers are formatted to have the same number of decimal places. You might want to take a look at `?biostats2SummaryFunctions` and  `?format_fixed` to customize their behaviour.




## Margins for percentage calculation

When crossing two factor variables, like `cyl` and `am`, percentages are calculated by row. You can use the `margin` argument to change this behavior.
 
```{r crosstable-percent-row}
crosstable(mtcars2, am, mpg, by=vs, funs=moystd, margin="row") %>% 
  as_flextable(keep_id=TRUE)
```
```{r crosstable-percent-column}
crosstable(mtcars2, am, mpg, by=vs, funs=moystd, margin="column") %>% 
  as_flextable(keep_id=TRUE)
```
```{r crosstable-percent-none}
crosstable(mtcars2, am, mpg, by=vs, funs=moystd, margin="none") %>% 
  as_flextable(keep_id=TRUE)
```
```{r crosstable-percent-both}
#I also could have used margin="all" here
crosstable(mtcars2, am, mpg, by=vs, funs=moystd, margin=c("column","row","cell")) %>% 
  as_flextable(keep_id=TRUE) 
```

Of course, if `by` is not set, `margin` will have no effect and percentages will be calculated in column.

Missing values are not taken into account when calculing percentage calculation. You can change this behaviour by using `tidyr::replace_na` or `forcats::fct_explicit_na` on your dataset before applying `crosstable`. 



## Totals

`crosstable` can also compute totals and display it in rows and/or columns with the `total` argument:

```{r crosstable-total-row}
crosstable(mtcars2, am, mpg, by=vs, funs=mean, total="row") %>% 
  as_flextable(keep_id=TRUE)
```
```{r crosstable-total-column}
#of course, total="column" only has an effect on categorical variables.
crosstable(mtcars2, am, mpg, by=vs, funs=mean, total="column") %>% 
  as_flextable(keep_id=TRUE)
```
```{r crosstable-total-both}
crosstable(mtcars2, am, mpg, by=vs, funs=mean, total="both") %>% 
  as_flextable(keep_id=TRUE)
```

Of course, if `by` is not set, `total` will always be calculated in column. 

Totals take missing values into account. Therefore, be aware that if argument `showNA="no"`, totals may be higher than the sum of the values inside the table.



## Tests

It is possible to perform statistical tests automatically. 

```{r crosstable-tests}
library(flextable)
crosstable(mtcars2, vs, qsec, by=am, funs=mean, test=TRUE) %>% 
  as_flextable(keep_id=TRUE)
```

Of course, this should only be done in an exploratory context, as it would cause extensive [alpha inflation](https://en.wikipedia.org/wiki/Multiple_comparisons_problem) otherwise.

Tests are chosen depending on the characteristics of the crossed variables (class, size, distribution, ...). See `?crosstable_test_args` for more details on the test choice algorithm.




## Effects

If the `by` variable has only 2 levels, it is also possible to automatically compute an effect. 

```{r crosstable-effect}
library(flextable)
crosstable(mtcars2, vs, qsec, by=am, funs=mean, effect=TRUE) %>% 
  as_flextable(keep_id=TRUE)
```

Type of effect (method, bootstrap, etc) are also chosen depending on the characteristics of the crossed variables (class, size, distribution, ...). See `?crosstable_effect_args` for more details on the effect choice algorithm.




## Miscellaneous

### Correlations 

If `by` refers to a numeric variable, correlation coefficients will be calculated. 

```{r crosstable-correlation}
library(survival)
crosstable(mtcars2, is.numeric, by=mpg) %>% 
  as_flextable(keep_id=TRUE)
```

Note that you can use the `cor_method` argument to choose which coefficient to calculate (`"pearson"`, `"kendall"`, or `"spearman"`).

### Survival data 

Crosstable is also able to describe survival data on specific times: 

```{r crosstable-survival}
library(survival)
aml$surv = Surv(aml$time, aml$status)
crosstable(aml, surv, by=x, times=c(0,15,30,150), followup=TRUE) %>% 
  as_flextable(keep_id=TRUE)
```

Using the formula interface, you can also declare the `Surv` object directly inside the `crosstable` function: `crosstable(aml, Surv(time, status) ~ x)`.



## Acknowledgement
`crosstable` is a rewrite of the awesome [`biostat2` package](https://github.com/eusebe/biostat2) written by David Hajage. The user interface is quite different but the concept is the same.

Thanks David!
