---
title: "Using `percent_pattern`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `percent_pattern`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=FALSE, 
  message=FALSE
)
options(width = 100)
```

In categorical variables, `crosstable()` offers a very versatile output using the `percent_pattern` argument.

This vignette will review the many things you can do using `percent_pattern`.


## Init

First, let's add some missing values to the `mtcars2` dataset and tweak some options:


```{r setup}
library(crosstable)
mtcars3 = mtcars2
mtcars3$cyl[1:5] = NA
mtcars3$vs[5:12] = NA

crosstable_options(
  percent_digits=0
)
```


## Default behaviour

By default, `crosstable()` will use `percent_pattern="{n} ({p_row})"`, so it outputs the size `n` along with the row's percentage `p_row`:

```{r}
crosstable(mtcars3, cyl, by=vs) %>% as_flextable()
```

## Allowed variables

Here is the list of all the internal variables you can use:

 * `n`, `n_row`, `n_col`, and `n_tot`: respectively the size of the cell, the row, the column, and the whole table.
 * `p_row`, `p_col`, and `p_tot`: respectively the proportion relative to the row, the column, and the whole table.
 * `p_tot_inf`, `p_tot_sup`, `p_row_inf`, `p_row_sup`, `p_col_inf`, `p_col_sup`: the confidence interval (calculated using [Wilson score](https://en.wikipedia.org/wiki/Binomial_proportion_confidence_interval#Wilson_score_interval)) for each of the proportions above. 
 
If you need it, you can also use any external variable defined outside of crosstable. 

Here is a simple example:
 
```{r}
crosstable(mtcars3, cyl, by=vs, 
           percent_pattern="N={n}, p={p_row} ({n}/{n_row})") %>% 
  as_flextable()
```
 
 
As you can see, these internal variables do not account for missing values (except for `n`, obviously).

This should make sense in most cases, but if it doesn't, you can use the following variables to account for NA explicitly:
 
 * `n_row_na`, `n_col_na`, `n_tot_na` 
 * `p_tot_na`, `p_row_na`, `p_col_na` 
 * `p_tot_na_inf`, `p_tot_na_sup`, `p_row_na_inf`, `p_row_na_sup`, `p_col_na_inf`, `p_col_na_sup` 

## Proportions in totals

As you may have noticed, totals are considered separately:

```{r}
crosstable(mtcars3, cyl, by=vs, total=TRUE, 
           percent_pattern="N={n}, p={p_row} ({n}/{n_row})") %>% 
  as_flextable()
```

Indeed, you cannot have the same pattern for totals, as the proportion relative to the row would not make sense in the context of the entire row itself.

To get control over the `percent_pattern` in totals, you have to pass a list with names `body`, `total_row`, `total_col`, and `total_all`:

```{r}
pp = list(body="N={n}, p={p_tot} ({n}/{n_tot})", 
          total_row="N={n} p=({p_col})", 
          total_col="{n}", total_all="Total={n_col}")
crosstable(mtcars3, cyl, by=vs, total=TRUE, 
           percent_pattern=pp) %>% 
  as_flextable()

```



## `get_percent_pattern()`

To easily get a `percent_pattern` list, you can use the `get_percent_pattern()` helper:

```{r}
get_percent_pattern("all")
get_percent_pattern("col", na=TRUE)
```

You can also set the result to a variable and modify its members at will. See `?get_percent_pattern` for more information.

## Ultimate example

Here is the ultimate example for `percent_pattern`. Give a close look to all possible values and you will surely find the one that you need.

```{r}
ULTIMATE_PATTERN=list(
  body="N={n}
        Cell: p = {p_tot} ({n}/{n_tot}) [{p_tot_inf}; {p_tot_sup}]
        Col: p = {p_col} ({n}/{n_col}) [{p_col_inf}; {p_col_sup}]
        Row: p = {p_row} ({n}/{n_row}) [{p_row_inf}; {p_row_sup}]
        
        Cell (NA): p = {p_tot_na} ({n}/{n_tot_na}) [{p_tot_na_inf}; {p_tot_na_sup}]
        Col (NA): p = {p_col_na} ({n}/{n_col_na}) [{p_col_na_inf}; {p_col_na_sup}]
        Row (NA): p = {p_row_na} ({n}/{n_row_na}) [{p_row_na_inf}; {p_row_na_sup}]",
  total_row="N={n}
             Row: p = {p_row} ({n}/{n_row}) [{p_row_inf}; {p_row_sup}]
             Row (NA): p = {p_row_na} ({n}/{n_row_na}) [{p_row_na_inf}; {p_row_na_sup}]",
  total_col="N={n}
             Col: p = {p_col} ({n}/{n_col}) [{p_col_inf}; {p_col_sup}]
             Col (NA): p = {p_col_na} ({n}/{n_col_na}) [{p_col_na_inf}; {p_col_na_sup}]",
  total_all="N={n}
             P: {p_col} [{p_col_inf}; {p_col_sup}]
             P (NA): {p_col} [{p_col_na_inf}; {p_col_na_sup}]"
)

crosstable(mtcars3, cyl, by=vs,
           percent_digits=0, total=TRUE, showNA="always",
           percent_pattern=ULTIMATE_PATTERN) %>% 
  as_flextable() %>% 
  flextable::theme_box()
```


Note that if you use `showNA="no"`
